Analyzing difficulty of Wordle using linguistic characteristics to determine

### 2022-05-07 outputs/paper/paper.Rmd
some minor changes for referencing and typo errors

---
title: Analyzing difficulty of Wordle using linguistic characteristics to determine
  average success of Twitter players
author: "Ivan Li"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: default
abstract: Using the Twitter API, data was gathered recording the public scores of
  Wordle players over one month. The frequency of each daily score was obtained and
  analyzed based on level of success, geographical location, and linguistic properties
  of the word. It was found that the linguistic properties of words such as commonality
  have impacts on player success. These results may have significance in literary
  and linguistic analysis.
thanks: 'Code and data are available at: https://github.com/Ivannoar/Twilight.'
toc: no
bibliography: references.bib
always_allow_html: yes
---

@@

library(tidyverse)
library(kableExtra)
library(countrycode)
library(chron)
library(ds4psy)
library(scales)
library(DiagrammeR)
library(knitr)
library(tinytex)
library(haven)

masterdata <- read_csv("masterdata.csv")
masterdata_clean <- read.csv("masterdata_clean.csv")

@@

# Introduction
(put citations later)
Wordle is an online word-guessing game bearing resemblance to historically popular word puzzles such as the New York Times crossword puzzle and the board game Mastermind. Players are given a five-letter word each day and are asked to guess the word within six tries. If the word is not guessed in earlier attempts, the letters for your word will become either grey, yellow, or green: grey indicating that the letter was not found anywhere in the word, yellow indicating the presence of the letter but in the wrong position, and green indicating both letter and position are correct. The goal is to guess five green letters and therefore the correct word. The game was created by Josh Wardle, a software engineer who created the game for himself and his partner to pass the time during the 2021 COVID-19 pandemic. The meteoric rise in popularity of his creation has been attributed to a feature where players can easily copy and paste their results to share with their friends and family. Wardle's game found international spread in late December of 2021 thanks to multiple factors, including the ease in sharing results and the daily nature of the game, as well as its simplicity. Wordle's popularity can be seen to its fullest extent on Twitter, where millions of users share their results every day by pasting their scores via the website's built-in feature.

This paper aims to analyze several metrics of Wordle's popularity on Twitter. As a result of millions of users sharing their scores each day, analysis can and has been done on variables including average score- a recent study found that Sweden was the country with the lowest average guesses with 3.72 average guesses out of a maximum of 6. Here, some key findings we wish to seek include attempting to model whether certain characteristics of a word correlate with its average difficulty, and the average score of Twitter users in relation to their geographical location, such as their country of residence. These metrics were not discussed deeply in prior papers and may be of interest to future puzzle makers, linguists, and game theorists.

Using the Twitter API, tweets pertaining to Wordle were gathered and aggregated into multiple datasets used for analysis. These tweets were gathered during intervals from April 4-10 2022, and April 17-23 2022). User scores were collected by searching for common templates that are used by the website domain to share scores in their tweets. Along with the score the used reached, the time of posting, geographical location, and other data provided by Twitter were also obtained. The results were turned into graphical data and analyzed according to our key missions.

@@

Finally, we conduct analysis on the words used in the game themselves. The following table shows the date the word was used, the average score users scored on it, the word of the day, the frequency of the word in texts according to the Corpus of Contemporary American English (COCA), and the number of orthographic neighbors to the word according to MCWord. Orthographic neighbors are defined as words which are the same length but differ by one letter (for example, scare and stare).

@@

In the time of writing of this paper, numerous reports and articles have been written about various aspects of Wordle, thanks to its meteoric rise in popularity since the beginning of 2022 as a result of the COVID-19 pandemic. This paper attempts to go a step further into analysis and cover the linguistic aspects of the popular word game and if any significant conclusions can be drawn from the results of the author's findings. The models constructed in this paper lead to both conclusions that are both reflected in the data and which are surprising at first glance but can be understood. It was the paper's hypothesis that the number of orthographic neighbours a word has would negatively impact the performance of the player; the failure rate and the number of guesses taken would both increase the similar a word was to others. This was concluded from the author's own experiences and the thought that words sharing high similarity would lead to many incorrect guesses used on the neighbours. However, the four models constructed show that while the less common a word is, the harder it is to guess, as predicted, the number of neighbours has a negative correlation with the average number of guesses as well as the failure rate of both types of players. Some explanations could include that having more neighbours means that the word's structure is easier to recall and thus provides the player with an early lead if guessed, or that the common structure of the word is correlated with words that players like to begin with as their first or second words, and therefore are easier to obtain hints on at the beginning of the game.

@@

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- The data was directly observable from each tweet and user, as tweets are sent as raw text and other details are directly obtainible from the timestamp and self-provided user information. 
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- The data was collected using the rtweet package found in R. These procedures were validated by manually checking that the program was gathering valid tweets using the status id and referencing of tweets and accounts. 
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- The sampling strategy was to take recent tweets matching the search criteria as of the time of request. Recent tweets are shown in order of time posted.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Only the author of the paper was involved in data collection.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- The data was collected over two timeframes- on April 10 and on April 23 2022. The timeframe roughly matches the creation timeframe of the data: The creation timeframes are roughly from April 4-10 and April 17-23 2022. These times are when daily tweets are collected.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- None
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- The data was collected from other sources, using the rtweet package.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- The individuals in question were not notified about the data collection.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- The individuals in question consented to the collection and use of their data in the Twitter Terms of Service which all users must agree to for site access. The link can be found here: https://twitter.com/en/tos

>>>>

---
title: Analyzing difficulty of Wordle using linguistic characteristics to determine
  average success of Twitter players
author: "Ivan Li"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: pdf_document
abstract: Using the Twitter API, data was gathered recording the public scores of
  Wordle players over one month. The frequency of each daily score was obtained and
  analyzed based on level of success, geographical location, and linguistic properties
  of the word. It was found that the linguistic properties of words such as commonality
  have impacts on player success. These results may have significance in literary
  and linguistic analysis.
thanks: 'Code and data are available at: https://github.com/Ivannoar/Twilight.'
toc: no
bibliography: references.bib
always_allow_html: yes
---

@@

library(tidyverse)
library(kableExtra)
library(countrycode)
library(chron)
library(ds4psy)
library(scales)
library(DiagrammeR)
library(knitr)
library(tinytex)
library(haven)
library(ggplot2)
masterdata <- read_csv("masterdata.csv")
masterdata_clean <- read.csv("masterdata_clean.csv")

@@

# Introduction

Wordle is an online word-guessing game bearing resemblance to historically popular word puzzles such as the New York Times crossword puzzle and the board game Mastermind. Players are given a five-letter word each day and are asked to guess the word within six tries. If the word is not guessed in earlier attempts, the letters for your word will become either grey, yellow, or green: grey indicating that the letter was not found anywhere in the word, yellow indicating the presence of the letter but in the wrong position, and green indicating both letter and position are correct. The goal is to guess five green letters and therefore the correct word. The game was created by Josh Wardle, a software engineer who created the game for himself and his partner to pass the time during the 2021 COVID-19 pandemic. The meteoric rise in popularity of his creation has been attributed to a feature where players can easily copy and paste their results to share with their friends and family. Wardle's game found international spread in late December of 2021 thanks to multiple factors, including the ease in sharing results and the daily nature of the game, as well as its simplicity. Wordle's popularity can be seen to its fullest extent on Twitter, where millions of users share their results every day by pasting their scores via the website's built-in feature.

This paper aims to analyze several metrics of Wordle's popularity on Twitter. As a result of millions of users sharing their scores each day, analysis can and has been done on variables including average score- a recent study found that Sweden was the country with the lowest average guesses with 3.72 average guesses out of a maximum of 6 [@citewordtips]. Here, some key findings we wish to seek include attempting to model whether certain characteristics of a word correlate with its average difficulty, and the average score of Twitter users in relation to their geographical location, such as their country of residence. These metrics were not discussed deeply in prior papers and may be of interest to future puzzle makers, linguists, and game theorists.

Using the Twitter API, tweets pertaining to Wordle were gathered and aggregated into multiple datasets used for analysis. These tweets were gathered during intervals from April 4-10, 2022, and April 17-23, 2022). User scores were collected by searching for common templates that are used by the website domain to share scores in their tweets. Along with the score the used reached, the time of posting, geographical location, and other data provided by Twitter were also obtained. The results were turned into graphical data and analyzed according to our key missions.

@@

Finally, we conduct analysis on the words used in the game themselves. The following table shows the date the word was used, the average score users scored on it, the word of the day, the frequency of the word in texts according to the Corpus of Contemporary American English (COCA) [@citecoca], and the number of orthographic neighbors to the word according to MCWord [@citemcword]. Orthographic neighbors are defined as words which are the same length but differ by one letter (for example, scare and stare).

@@

In the time of writing of this paper, numerous reports and articles have been written about various aspects of Wordle, thanks to its meteoric rise in popularity since late 2021 as a result of the COVID-19 pandemic. This paper attempts to go a step further into analysis and cover the linguistic aspects of the popular word game and if any significant conclusions can be drawn from the results of the author's findings. The models constructed in this paper lead to both conclusions that are both reflected in the data and which are surprising at first glance but can be understood. It was the paper's hypothesis that the number of orthographic neighbours a word has would negatively impact the performance of the player; the failure rate and the number of guesses taken would both increase the similar a word was to others. This was concluded from the author's own experiences and the thought that words sharing high similarity would lead to many incorrect guesses used on the neighbours. However, the four models constructed show that while the less common a word is, the harder it is to guess, as predicted, the number of neighbours has a negative correlation with the average number of guesses as well as the failure rate of both types of players. Some explanations could include that having more neighbours means that the word's structure is easier to recall and thus provides the player with an early lead if guessed, or that the common structure of the word is correlated with words that players like to begin with as their first or second words, and therefore are easier to obtain hints on at the beginning of the game.

@@

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- The data was directly observable from each tweet and user, as tweets are sent as raw text and other details are directly obtainible from the timestamp and self-provided user information. 
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- The data was collected using the rtweet package found in R. These procedures were validated by manually checking that the program was gathering valid tweets using the status id and referencing of tweets and accounts. 
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- The sampling strategy was to take recent tweets matching the search criteria as of the time of request. Recent tweets are shown in order of time posted.
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Only the author of the paper was involved in data collection.
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- The data was collected over two timeframes- on April 10 and on April 23, 2022. The timeframe roughly matches the creation timeframe of the data: The creation timeframes are roughly from April 4-10 and April 17-23, 2022. These times are when daily tweets are collected.
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- None
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- The data was collected from other sources, using the rtweet package.
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- The individuals in question were not notified about the data collection.
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- The individuals in question consented to the collection and use of their data in the Twitter Terms of Service which all users must agree to for site access. The link can be found here: https://twitter.com/en/tos
